---
title: "Iteration Notes"
subtitle: "R for Data Science"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, dpi=300)

library(tidyverse)
```

## Iteration
Iteration is useful when you want to perform the same action many times. The
most common type of loop is called the `for` loop.

## Why haven't we used loops?
```{r no-loops-yet}
x <- c(100, 200, 300)
y <- 2*x
y
```

::: notes
We haven't used loops yet since most of the operations we've seen are already 
*vectorized*. In the example above, we are able to quickly calculate `y` without
iterating over every value in `x`
:::

## Other languages teach this way first
```{r other-language-loops}
y <- vector("double", length(x))

for (i in 1:length(x)) {
  y[i] <- 2*x[i]
}

y
```

::: notes
Most programming books on other languages introduce loops early on. This is how
they would have solved the previous problem. You iterate over every value in x
and perform a computation on it (in this case, multiplying by two). The result
of that computation is saved in the new appropriate place in the output.

There are three parts to loop: 
  1. initialize the output 
  2. iterate over a sequence
  3. perform a computation
:::

## For Loop: Initialize Output
```{r init-output}
y <- vector("double", length(x))

for (i in 1:length(x)) {
  y[i] <- 2*x[i]
}
```

::: notes
The first thing we do is create an empty place to store our output. This output
will be the same length as the number of iterations we plan on doing.
:::

## Init Output 2
`y <- vector("double", length(x))`

```{r init-output2, echo=FALSE}
df <- tibble(x = c(100, 200, 300), y = c("", "", ""))
knitr::kable(df)
```

## For Loop: Sequence
```{r init-sequence}
y <- vector("double", length(x))

for (i in 1:length(x)) {
  y[i] <- 2*x[i]
}
```
::: notes
The second thing we do is define the sequence of values we will be iterating over.
In this case, we are iterating over the values between 1 and length(x)
:::

## For Loop: Sequence 2
```{r init-sequence2}
for (i in 1:5) {
  print(i)
}

for (i in 5:1) {
  print(i)
}
```

## For Loop: Body
```{r init-body}
y <- vector("double", length(x))

for (i in 1:length(x)) {
  y[i] <- 2*x[i]
}
```

::: notes
The final thing we do is define the body of the computation. Note we are using the
`[` accessor to get the i-th element of `x` and `y`.
:::

## Calculate the mean of every column
How do I calculate the mean of every column in `mtcars`?

```{r quiz-1, echo=FALSE}
knitr::kable(head(mtcars, 4))
```

## 1. Define your output
```{r mean-output}
n <- ncol(mtcars)
x <- vector("double", n)
```

## 2. Define your iteration sequence
```{r mean-sequence, eval=FALSE}
1:ncol(mtcars)
```

## 3. Define computation body
```{r mean-body, eval=FALSE}
x[i] <- mean(mtcars[[i]])
```

## All together now
```{r mean-for-loop}
n <- ncol(mtcars)
x <- vector("double", n)

for (i in 1:n) {
  x[i] <- mean(mtcars[[i]])
}

x
```

# For Loop Tips

## Tip 1: Initialize Your Output
You technically don't have to initialize your output. You can grow your output
vector with each iteration. Don't do this. Be better.

```{r growing-output}
y <- NA

for (i in 1:4) {
  y[i] <- i^2
}

y
```

::: notes
This is because R has to figure out each iteration how much more memory to allocate
for your object. It's a lot more efficient to allocate it all at once in the beginning.
:::

## Tip 2: Use seq_* functions

```{r seq-problem-along}
x <- vector("integer", 0)
for (i in 1:length(x)) {
  print(i)
}
```

```{r seq-better-along}
x <- vector("integer", 0)
for (i in seq_along(x)) {
  print(i)
}
```

```{r seq-problem-len}
n <- 0
for (i in 1:n) {
  print(i)
}
```

```{r seq-better-len}
n <- 0
for (i in seq_len(n)) {
  print(i)
}
```

# More For Loop Uses
1. Iterating over an object and creating a new one
2. Modifying the original object
3. Looping over names or values
4. Outputs of unknown length
5. Sequences of unknown length



## So, when do I use loops?
If an iteration depends on previous iterations, you have to use a loop. However,
many times that is not the case and you are better off using `map()` or `apply*()`